Selective Sequential Model Selection

### 2015-12-09 paper.tex
Fixed bib file

\begin{abstract}
  Many model selection algorithms produce a path of fits specifying a sequence of increasingly complex models. Given such a sequence and the data used to produce them, we consider the problem of choosing the least complex model that is not falsified by the data. Extending the selected-model tests of \citet{fithian2014optimal}, we construct $p$-values for each step in the path which account for the adaptive selection of the model path using the data. In the case of linear regression, we propose two specific tests, the max-$t$ test for forward stepwise regression (generalizing a proposal of \citet{buja2014}), and the next-entry test for the lasso. These tests improve on the power of the saturated-model test of \citet{taylor2014exact}, sometimes dramatically. In addition, our framework extends beyond linear regression to a much more general class of parametric and nonparametric model selection problems.

To select a model, we can feed our single-step $p$-values as inputs into sequential stopping rules such as those proposed by \citet{gsell2013sequential} and \citet{li2015accumulation}, achieving control of the familywise error rate or false discovery rate (FDR) as desired. The FDR-controlling rules require the null $p$-values to be independent of each other and of the non-null $p$-values, a condition not satisfied by the saturated-model $p$-values of \citet{taylor2014exact}. We derive intuitive and general sufficient conditions for independence, and show that our proposed constructions yield independent $p$-values.
\end{abstract}

\section{Introduction}
\label{sec:introduction}
Many model selection procedures produce a sequence of increasingly complex models, leaving the data analyst to choose among them. 
Given such a path, we consider the problem of choosing the simplest model in the path that is not falsified by the available data. Examples of path algorithms include forward stepwise linear regression, least angle regression (LAR) and the ever-active path in lasso ($\ell_1$-regularized) regression. \citet{taylor2014exact} study methods for generating exact $p$-values at each step of these path algorithms, and their methods provide a starting point for our proposals. Other related works include \citet{loftus2014significance}, who describe $p$-values for path algorithms that add groups of variables (instead of individual variables) at each step, and \citet{choi2014selecting}, who describe $p$-values for steps of principle components analysis (each step marking the estimation of a principle component direction).

We consider the following workflow: to select a model, we compute a set of sequential $p$-values at each step of the model path, and then feed them into a stopping rule that is guaranteed to control the false discovery rate (FDR) \citep{benjamini1995controlling}, familywise error rate (FWER), or a similar quantity. Recently \citet{gsell2013sequential} and \citet{li2015accumulation} proposed sequential stopping rules of this kind. Both sets of rules require that once we have reached the first correct model in our path, the $p$-values in subsequent steps are uniform and independent. While this is not true of the $p$-values constructed in \citet{taylor2014exact}, in this paper we develop a theoretical framework for constructing sequential $p$-values satisfying these properties, and give explicit constructions.

@@

In this article we will concern ourselves primarily with testing the complete null, reflecting our stated aim of {\em choosing the least complex model that is consistent with the data}. As we discuss further in Section~\ref{sec:selective-reg}, the saturated-model tests of \citet{taylor2014exact} are valid selective tests of $H_k^{\text{inc}}$. The advantage of these tests is that they are highly computationally efficient (they do not require sampling). But, unfortunately, they also carry a number of drawbacks: they require us to assume $\sigma^2$ is known, can result in a large reduction in power, create dependence between $p$-values at different steps, and are difficult to generalize beyond the case of linear regression.

@@

As we have shown, the selective max-$t$ test is equivalent to the selective greedy likelihood test in the case of linear regression. In the next section, we will compare and contrast the max-$t$ and next-entry, and other {\em selected-model tests}, with the {\em saturated-model tests} proposed by \citet{taylor2014exact} and others.

@@

\subsection{Saturated-Model Tests}
Many conditional selective tests for linear regression use the saturated-model framework. See for example \citet{taylor2013tests}, \citet{taylor2014exact}, \citet{lee2013exact}, and \citet{loftus2014significance}. Define the saturated model as
\[
M_{\text{sat}}:\; Y \sim \cN(\mu, \sigma^2I_n),
\]
so named because there is a mean parameter for every observation. Because the parameter $\mu$ has the same dimension as the data $Y$, we must assume that $\sigma^2$ is known. If $\sigma^2$ is unknown, we can estimate $\sigma^2$ and apply the test while plugging in the estimated value; this is what we have done with the diabetes data to compute the $p$-values in Table~\ref{tab:diab}.  \citet{tibshirani2015uniform} discuss this and other strategies for unknown $\sigma^2$.

@@

For a completely general selection algorithm, we might expect severe computational difficulties when we attempt to condition on the null model at stage $k$, $M_{k-1}(Y)$. For example, in forward stepwise regression, $M_{k-1}$ is the cumulative result of $k-1$ steps of greedy variable selection. At each step $s<k$, the selection of variable $j_s$ is a result of comparing its $t$-statistic with the $t$-statistics of the other $p-k$ inactive variables. Based on this logic, we might expect to accumulate at least $O(p)$ constraints at each step, resulting in $O(kp)$ constraints after $k$ steps \citep{taylor2014exact}.

>>>>

\begin{abstract}
  Many model selection algorithms produce a path of fits specifying a sequence of increasingly complex models. Given such a sequence and the data used to produce them, we consider the problem of choosing the least complex model that is not falsified by the data. Extending the selected-model tests of \citet{fithian2014optimal}, we construct $p$-values for each step in the path which account for the adaptive selection of the model path using the data. In the case of linear regression, we propose two specific tests, the max-$t$ test for forward stepwise regression (generalizing a proposal of \citet{buja2014}), and the next-entry test for the lasso. These tests improve on the power of the saturated-model test of \citet{tibshirani2014exact}, sometimes dramatically. In addition, our framework extends beyond linear regression to a much more general class of parametric and nonparametric model selection problems.

To select a model, we can feed our single-step $p$-values as inputs into sequential stopping rules such as those proposed by \citet{gsell2013sequential} and \citet{li2015accumulation}, achieving control of the familywise error rate or false discovery rate (FDR) as desired. The FDR-controlling rules require the null $p$-values to be independent of each other and of the non-null $p$-values, a condition not satisfied by the saturated-model $p$-values of \citet{tibshirani2014exact}. We derive intuitive and general sufficient conditions for independence, and show that our proposed constructions yield independent $p$-values.
\end{abstract}

\section{Introduction}
\label{sec:introduction}
Many model selection procedures produce a sequence of increasingly complex models, leaving the data analyst to choose among them. 
Given such a path, we consider the problem of choosing the simplest model in the path that is not falsified by the available data. Examples of path algorithms include forward stepwise linear regression, least angle regression (LAR) and the ever-active path in lasso ($\ell_1$-regularized) regression. \citet{tibshirani2014exact} study methods for generating exact $p$-values at each step of these path algorithms, and their methods provide a starting point for our proposals. Other related works include \citet{loftus2014significance}, who describe $p$-values for path algorithms that add groups of variables (instead of individual variables) at each step, and \citet{choi2014selecting}, who describe $p$-values for steps of principle components analysis (each step marking the estimation of a principle component direction).

We consider the following workflow: to select a model, we compute a set of sequential $p$-values at each step of the model path, and then feed them into a stopping rule that is guaranteed to control the false discovery rate (FDR) \citep{benjamini1995controlling}, familywise error rate (FWER), or a similar quantity. Recently \citet{gsell2013sequential} and \citet{li2015accumulation} proposed sequential stopping rules of this kind. Both sets of rules require that once we have reached the first correct model in our path, the $p$-values in subsequent steps are uniform and independent. While this is not true of the $p$-values constructed in \citet{tibshirani2014exact}, in this paper we develop a theoretical framework for constructing sequential $p$-values satisfying these properties, and give explicit constructions.

@@

In this article we will concern ourselves primarily with testing the complete null, reflecting our stated aim of {\em choosing the least complex model that is consistent with the data}. As we discuss further in Section~\ref{sec:selective-reg}, the saturated-model tests of \citet{tibshirani2014exact} are valid selective tests of $H_k^{\text{inc}}$. The advantage of these tests is that they are highly computationally efficient (they do not require sampling). But, unfortunately, they also carry a number of drawbacks: they require us to assume $\sigma^2$ is known, can result in a large reduction in power, create dependence between $p$-values at different steps, and are difficult to generalize beyond the case of linear regression.

@@

As we have shown, the selective max-$t$ test is equivalent to the selective greedy likelihood test in the case of linear regression. In the next section, we will compare and contrast the max-$t$ and next-entry, and other {\em selected-model tests}, with the {\em saturated-model tests} proposed by \citet{tibshirani2014exact} and others.

@@

\subsection{Saturated-Model Tests}
Many conditional selective tests for linear regression use the saturated-model framework. See for example \citet{taylor2013tests}, \citet{tibshirani2014exact}, \citet{lee2013exact}, and \citet{loftus2014significance}. Define the saturated model as
\[
M_{\text{sat}}:\; Y \sim \cN(\mu, \sigma^2I_n),
\]
so named because there is a mean parameter for every observation. Because the parameter $\mu$ has the same dimension as the data $Y$, we must assume that $\sigma^2$ is known. If $\sigma^2$ is unknown, we can estimate $\sigma^2$ and apply the test while plugging in the estimated value; this is what we have done with the diabetes data to compute the $p$-values in Table~\ref{tab:diab}.  \citet{tibshirani2015uniform} discuss this and other strategies for unknown $\sigma^2$.

@@

For a completely general selection algorithm, we might expect severe computational difficulties when we attempt to condition on the null model at stage $k$, $M_{k-1}(Y)$. For example, in forward stepwise regression, $M_{k-1}$ is the cumulative result of $k-1$ steps of greedy variable selection. At each step $s<k$, the selection of variable $j_s$ is a result of comparing its $t$-statistic with the $t$-statistics of the other $p-k$ inactive variables. Based on this logic, we might expect to accumulate at least $O(p)$ constraints at each step, resulting in $O(kp)$ constraints after $k$ steps \citep{tibshirani2014exact}.
